---
title: "<br><br>Pediatric mental health emergencies across three waves of the COVID-19 pandemic in New York City"
output:
  html_document:
    css: styles.css
    includes:
      before_body: footer.html
---

<br><br>

### Abstract

**Background**
The mental health of children has been adversely affected by the COVID-19 pandemic. In New York City (NYC), an epicenter of the pandemic, three distinct waves in cases each posed unique and growing psychosocial challenges for children.

**Objective**
We aimed to describe trends in patient volume, sociodemographics, clinical diagnoses, and disposition of mental health emergencies (MHEs) in children in NYC during each wave of the COVID-19 pandemic.

**Methods**
We conducted a cross-sectional analysis of pediatric emergency department (PED) visits using electronic health record (EHR) data from the INSIGHT Clinical Research Network, a large multicenter database comprised from five major institutions in NYC. We examined four study periods (dates determined by NYC Department of Health data trends and study investigator consensus): pre-pandemic (1/1/16-2/28/20), wave 1 (3/1-7/31/20), wave 2 (8/1-6/30/21),wave 3 (7/1-9/20/21). We described changes in the number of MHEs per 100,000 PED visits, sociodemographic characteristics (including social deprivation index (SDI), race, and ethnicity), clinical diagnoses, and disposition during the pandemic waves compared to pre-pandemic. We used descriptive statistics and unadjusted, two-tailed, two-sample tests for equality of proportions.

**Results**
We included 18,652 patients with MHEs. The proportion of MHEs in children presenting to the PED has been increasing since 2016, with a large rise during the COVID-19 pandemic, especially during the 2nd wave (Figure 1). The proportion of female, White, and adolescent patients increased throughout the pandemic. Median SDI was lower compared to pre-pandemic. The proportions of MHEs with anxiety and substance abuse diagnoses were highest in the 1st wave. The proportions of MHEs with depression, eating disorders, and suicide/self-injury diagnoses were highest in the 2nd wave. The proportion of MHEs with autism spectrum disorders was highest during the 3rd wave (Table 1). A greater proportion of pediatric MHEs resulted in hospitalization during the pandemic compared to pre-pandemic (2.4% vs 1.2%; difference 1.2%; 95% CI: 0.8%-1.7%). Descriptively, a higher proportion of Black and Hispanic children returned to the PED within 30 days during the pandemic compared to pre-pandemic (Table 2).

**Conclusions**
The proportion of MHEs in children presenting to the PED in NYC rose throughout the COVID-19 pandemic and was highest during the 2nd wave. Trends over time differed by mental health diagnosis. Increased admissions may signify higher MHE acuity. Sociodemographic differences in PED visits may represent inequitable access to mental healthcare.

<br> 

### Tables {.tabset .tabset-fade}

#### Table 1

```{=html}
```{r, echo=FALSE, results='asis'}
xfun::file_string('files/tbl_1_2022-03-16.html')
```
```

#### Table 2

```{=html}
```{r, echo=FALSE, results='asis'}
xfun::file_string('files/tbl_2.html')
```
```
<!-- <a href="https://simmwill.github.io/files/tbl_1.html" style="color:#000000; text-decoration: none">Click here to see Table 1.</a> -->

<!-- <a href="https://simmwill.github.io/files/tbl_2.html" style="color:#000000; text-decoration: none">Click here to see Table 2.</a> -->
